---
title: "03b_analyze_DTM"
author: "Erin Becker"
date: "October 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "~/Box Sync/digital-humanities-phylogenetics/")
```

## Analyze the Document Term Matrix
This continues our analysis of document similarity based on shared entries.

Note: The "EXTRA" tag indicates a hidden code chunk illustrating an important point about
the data that is tangential to the main thread of the analysis. You can run these
code chunks by setting eval=TRUE and see the code itself by setting echo=TRUE.

```{r}

#### Setup
## If you haven't already, install and load the required packages. 

# install.packages(c("ggplot2", "ggdendro", "reshape", "scales", "RColorBrewer", "gtools", "reshape2"))

#source("http://bioconductor.org/biocLite.R")
#biocLite("BiocInstaller")
#biocLite("phangorn")

library(ggplot2)
library(ggdendro)
library(reshape)
library(reshape2)
library(scales)
library(RColorBrewer)
library(gtools)
library(phangorn)
library(phytools)
```

Load in the document term matrix (DTM) that was generated previously.

```{r}
dtm_df = read.csv("data/pass/Q39_par_dtm.csv", 
                  stringsAsFactors = F, row.names = 1) 
```

### Check how R parsed the DTM
R doesn't allow variable names to start with a number, 
so entries starting with a number had "X" added to the beggining of the entry name.
R doesn't allow parentheses or brackets in variable names, 
so entries like "1(ban₂)[na]na" parsed as "X1.ban...na.na"

```{r}
head(dtm_df[,1:3])
```

In a later cell, we will sum the columns and rows to get an idea of the density of the matrix. In order to avoid iteratively summing the summary column and summary row, we need to set variables specifying the columns and rows we want to sum. 

```{r}
cols_to_sum = ncol(dtm_df)
rows_to_sum = nrow(dtm_df)

dim(dtm_df) 
```

### Check density of DTM

Two attributes of the DTM which are of interest to us are:
  
1) Distribution of document lengths (number of entries per document).  
2) Distribution of entry frequency (number of times each entry appears across the corpus). Note that an entry may appear multiple times in a single document.

We will look at these two distributions below.

Note that entries may appear multiple times in a single document (EXTRA).

```{r EXTRA, eval=FALSE, echo=FALSE}
# Output shows each entry that appears more than once within a document, along with the 
# maximum number of times that it appears within a document.

for(x in colnames(dtm_df)) {
  
  num_appearances = unname(unlist(unique(dtm_df[x])))
  if(any(num_appearances > 1)) print(paste(x, max(num_appearances)))
  
}
```

We next add a column showing the number of entries in each document, a row
showing the number of occurences of each entry, and set the intersection of these
two sumamries to NA.

```{r}
dtm_df$num_entries = rowSums(dtm_df[1:cols_to_sum])
dtm_df["num_occurrences",] = colSums(dtm_df[1:rows_to_sum,])
dtm_df["num_occurrences","num_entries"] = NA 
```

We can see the distribution of document length by looking at a histogram.

```{r}
hist(dtm_df$num_entries, breaks = 30, main = "Document length", 
     ylab = "# documents", xlab = "# entries")
```

There are `r length(which(dtm_df$num_entries >= 10))` documents with 10 or more entries and `r length(which(dtm_df$num_entries >= 100))` documents with 100 or more entries.

We see that the majority of documents are very short. However, there are a decent number of documents with a substantial number of entries (summarized above). 

## Look at document length
Which documents are the longest?

```{r}
dtm_df = dtm_df[order(dtm_df$num_entries, decreasing = TRUE),]
head(dtm_df["num_entries"])

num_occurrences = unlist(dtm_df["num_occurrences",])
hist(num_occurrences, breaks = 30, main = "Entry frequency", 
     ylab = "# entries", xlab = "# occurrences")
hist(num_occurrences, breaks = 30, ylim = c(0, 100), 
     main = "Entry frequency (zoomed in)", 
     ylab = "# entries", xlab = "# occurrences")
```
We then look at the distribution of entries across the corpus.

``` {r}
rare = round(length(which(dtm_df["num_occurrences",] <=2))/cols_to_sum*100,2)
common_entries = colnames(dtm_df[which(dtm_df["num_occurrences",] >=10)])
most_common = max(dtm_df["num_occurrences",], na.rm = TRUE)
most_common_entry = colnames(dtm_df[which(dtm_df["num_occurrences",] == most_common)])
```

`r rare`% of entries appear only once or twice across the corpus. `r length(common_entries)` entries occur 10 or more times within the corpus, including one that occurs `r most_common` times within the `r rows_to_sum` documents. The most common entry is `r most_common_entry`.

Below we inspect the most common entries (those that appear at least ten times across the corpus) to see if they make sense.

## Look at some of the most common entries

```{r}
common_entries
```

As seen from the output above, many of the most common entries include
either unlematizable or illegible words (represented as x.na.na or x.x.na.na). 
These are not particularly informative. We want to remove these entries from the analysis. Also remove the entry "unknown".

```{r}
entries_with_na = grep("na.na", colnames(dtm_df))

if(length(entries_with_na) > 0) {
  dtm_df = dtm_df[,-entries_with_na]
  dtm_df$unknown = NULL
}
```

Look again at entry frequency after removing these entries.

Note that we must re-create the summary column and row after excluding these entries.

```{r}
# Remove summary column and row
dtm_df$num_entries = NULL
dtm_df = dtm_df[-which(rownames(dtm_df) == "num_occurrences"),]

#recreate summary column and row
cols_to_sum = ncol(dtm_df)
rows_to_sum = nrow(dtm_df)

dtm_df$num_entries = rowSums(dtm_df[1:cols_to_sum])
dtm_df["num_occurrences",] = colSums(dtm_df[1:rows_to_sum,])
dtm_df["num_occurrences","num_entries"] = NA 
```
```{r}
num_occurrences = unlist(dtm_df["num_occurrences",])
hist(num_occurrences, breaks = 30, main = "Entry frequency", 
     ylab = "# entries", xlab = "# occurrences")
hist(num_occurrences, breaks = 30, ylim = c(0, 100), 
     main = "Entry frequency (zoomed in)", 
     ylab = "# entries", xlab = "# occurrences")

rare = round(length(which(dtm_df["num_occurrences",] <=2))/cols_to_sum*100,2)
common_entries = colnames(dtm_df[which(dtm_df["num_occurrences",] >=10)])
most_common = max(dtm_df["num_occurrences",], na.rm = TRUE)
most_common_entry = colnames(dtm_df[which(dtm_df["num_occurrences",] == most_common)])
```

`r rare`% of entries appear only once or twice across the corpus. `r length(common_entries)` entries occur 10 or more times within the corpus, including one that occurs `r most_common` times within the `r rows_to_sum` documents. The most common entry is `r most_common_entry`. Now some of the most common entries are: 

```{r}
common_entries
```

Next we will reorganize our data to enable pretty and informative plotting of entry distribution across documents and overall frequency.

First, we reorder the columns in our dataframe from alphabetical to
sorting by frequency so that the most common entries are clustered together for ease of visualization.

To do this, we first convert the frequency counts to presence/absence (binary), then reorder the columns by column sums.

```{r}
# Convert to binary presence/absence information
dtm_df = as.matrix((dtm_df > 0) + 0)

dtm_df = dtm_df[,order(colSums(dtm_df), decreasing = TRUE)]
dtm_df = as.data.frame(dtm_df)
```

We then recalculate number of occurrences
and number of entries. These are now binary, so they
represent:
  
- num_occurrences = number of documents in which an entry appears  
- num_entries = number of unique entries in a document (excluding "na.na" and "unknown" entries)

``` {r}
dtm_df$num_entries = rowSums(dtm_df[1:cols_to_sum])
dtm_df["num_occurrences",] = colSums(dtm_df[1:rows_to_sum,])
dtm_df["num_occurrences","num_entries"] = NA 

num_occurrences = unlist(dtm_df["num_occurrences",])
most_frequent = max(num_occurrences, na.rm = TRUE)
most_frequent_entry = colnames(dtm_df[which(dtm_df["num_occurrences",] == most_frequent)])
```

`r table(num_occurrences)[1]` entries appear in only one document. The entry that appears in the most documents is `r most_frequent_entry`.

```{r}
# table(num_occurrences, useNA = "ifany")

num_occurrences = unlist(dtm_df["num_occurrences",])
hist(num_occurrences, breaks = max(num_occurrences, na.rm = TRUE), main = "Entry frequency", 
     ylab = "# entries", xlab = "# occurrences")
```

These entries appear in at least 10 different documents:
```{r}
colnames(dtm_df)[which(dtm_df["num_occurrences",] >=10)]
```

Our dataset is built around lists of trees and wooden objects (composite text Q000039). For sanity testing our analyses, we've included some outlier texts 
which are not part of the lists of trees and wooden objects document group. 
One of the outlier text is Q000001, which is a composite text of lists of animals.
Before using this text as an outlier, we want to make sure that it doesn't share
many similarities with Q000039. We find very few shared entries between Q000039 and Q000001. (EXTRA)

``` {r echo = FALSE, eval = FALSE}
# Which entries appear in both Q000001 and Q000039?
shared_entries = which(dtm_df["Q000001",] == 1 & dtm_df["Q000039",] == 1)
dtm_df[c("Q000001", "Q000039"),shared_entries]

# Do these entries also appear in other documents?
docs_w_shared_entries = c()

for(x in rownames(dtm_df)) {
  if(any(dtm_df[x,shared_entries] > 0)) {
    docs_w_shared_entries = c(docs_w_shared_entries,x)
  }}

dtm_df[docs_w_shared_entries,shared_entries]
```

### Remove rare entries
A large number of entries appear only in one document in the corpus. We remove these from the remainder of the analysis. However, we want to keep all entries from our outlier texts. After removing rare entries, the dimensions of our dtm are:

```{r}
# Keep all entries which appear in one of the outlier texts.
outlier_docs = c("Q000001", "P250736", "num_occurrences")
outlier_doc_data = dtm_df[outlier_docs,]
keep1 = which(outlier_doc_data["num_occurrences",] == 1)

# Also keep entries which appear in more than one document in the corpus.
keep2 = which(dtm_df["num_occurrences",] > 1)

all_keep = unique(c(keep1, keep2))

dtm_df = dtm_df[,all_keep]

dim(dtm_df)
```

Our next step is to compare this presence/absence distribution of entries across documents to understand which documents are most similar to one another at the entry level. First we will extract only those texts with at least ten entries to see if patterns make sense before analyzing the shorter texts. After removing short texts, the dimmensions of our dataframe are: 

``` {r}
# add back in num_entries column
dtm_df$num_entries = rowSums(dtm_df[1:cols_to_sum])

dtm_df = dtm_df[which(dtm_df$num_entries >= 10),]

#add document names as row names
#dtm_df$document = rownames(dtm_df)

# remove num_occurances row
dtm_df = dtm_df[-which(rownames(dtm_df) == "num_occurrences"),]

# remove num_entries column
dtm_df$num_entries = NULL

dim(dtm_df)
```

### Write out cleaned dtm

After these cleaning steps, we now have the DTM that we will use for our remaining analyses. We will write out a copy of this dataframe so that we can pass it in as an 
object to other scripts. 

```{r echo = TRUE, eval = FALSE}
#add document names as row names
dtm_df$document = rownames(dtm_df)

write.csv(dtm_df, "data/dtm_data/dtm_df_clean.csv", quote = FALSE)
```
