---
title: "04_define_sections"
author: "Erin Becker"
date: "November 1, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "~/Box Sync/digital-humanities-phylogenetics/")
```

## Define Sections based on composite text

The first step in analyzing order of entries within a section is defining what constitutes a section. We've discussed three methods for defining sections.

1) Expert definition.
2) Automated based on composite text (eg. Q000039).
3) Automated based on entry proximity.

This document implements the second method (automated based on composite text).

A composite text is read in and breakpoints in the text are determined based on similar words (either in base or lemma). Entries between those breakpoints are assumed to belong to
the same section. 

Based on the following (artificial) Q text, this method would ideally lead to three sections. Lines 1-4 (related to palm), lines 5-8 (related to poplar), and lines 9-14 (related to tree).

1) ŋešnimbar[palm]N
2) ŋešnimbar[palm]N sux:suhuš[offshoot]N
3) deg[collect]V/t sux:ŋešnimbar[palm]N
4) niŋkiluh[broom]N sux:ŋešnimbar[palm]N
5) asal[poplar]N
6) asal[poplar]N sux:kur[mountain]N
7) asal[poplar]N sux:dug[good]V/i
8) numun[seed]N sux:asal[poplar]N
9) ilur[tree]N
10) ad[bush]N
11) kišig[acacia]N
12) kišighar[tree]N
13) samazum[tree]N
14) peškal[tree]N

Note that in our current implementation, the split is actually entries 1-4, entries 6-8, and entries 11-14. Entries 9 and 10 are not grouped with any section. This will be explained further in a later note.

### Implementation strategy
Our strategy for implementing this method is the following:

1) Calculate the degree of similarity between any two adjacent entries in the source/composite document. This similarity can take two independent forms:
  a) shared character strings in the Sumerian (citation form) or the English (guideword)
  b) shared entire words (citation forms and/or guidewords).
2) Determine appropriate cutoff values for similarity that distinguish between sections.
3) Automatically assign a name to each section.
4) Output a file containing the entries present in each section.

We can then use these section definitions to analyze similarities and differences between texts in terms of both section presence / absence patterns and section order.

```{r}

### Setup
## If you haven't already, install and load the required packages. 

# install.packages(c("ggplot2", "ggdendro", "reshape", "scales", "dplyr"))
# 
# source("http://bioconductor.org/biocLite.R")
# biocLite("BiocInstaller")
# biocLite("Rlibstree")

library(ggplot2)
library(ggdendro)
library(reshape)
library(scales)
library(Rlibstree)
library(dplyr)
```

## Step 1: Calculate similarity between adjacent entries.

This step relies on defining functions to process and clean entry strings. The following entry will be used to demonstrate these functions:
```{r}
entry = "e[house]N_usan[whip]N_gigir[chariot]N"
print(entry)
```

### Extract whole words

This method for defining sections relies in part on looking for words shared between entries. These shared words can be citation forms (Sumerian) and/or guidewords (English). The functions below enable us to extract guidewords and citation forms from entries.

```{r echo = TRUE}
get_guidewords = function(line) {
  # extract all guidewords for an entry into character vector
  line = unlist(strsplit(line, "_"))
  line = gsub(".*\\[", "", line)
  line = gsub("\\].*", "", line)
  line = gsub("~", "", line)
  guidewords = line
  guidewords = guidewords[which(guidewords != "na")]
  guidewords
}

get_citation_forms = function(line) {
  # extract all citation forms for an entry into character vector
  line = unlist(strsplit(line, "_"))
  line = gsub("(.*)\\[.*", "\\1", line)
  citation_form = line
  citation_form = citation_form[which(citation_form != "na")]
  citation_form
}
```

#### Example to illustrate functions
```{r echo = TRUE}
print(entry)
get_guidewords(entry)
get_citation_forms(entry)
```

Entries may also share partial words (either in the citation form or the guideword) (e.g. "kišig[acacia]n" and "kišighar[tree]n" share "kišig", which would not be found using the whole word approach). We need to find these partial word matches in both the Sumerian (citation forms) and the English (guidewords). The function below finds the longest common substring shared between any two citation forms or two guidewords for an entry pair.

```{r}
get_longest_kmer_from_set_of_words = function(entry1, entry2, fun) {
  # extract word sets from each entry
  word_set_x = fun(entry1)
  word_set_y = fun(entry2)
  
  # initialize empty vector to store results
  longest = c() 
  
  # loop through each word pair and get longest common substring
  for (word_x in word_set_x) {
    for (word_y in word_set_y) {
      substring = getLongestCommonSubstring(c(word_x, word_y))
      longest = c(longest, gsub("[\x80-\xFF]", "", substring)) # removes random multibyte characters sometimes added by the getLongestCommonSubstring function
    }}
  
  if (length(longest) == 0) {
    kmer = NA
    k = 0 }
  
  else {
  # get longest common substring for any of the possible pairs
  # if there is more than one longest substring of the same length, take the first
  k = max(nchar(longest))[1]
  kmer = longest[which(nchar(longest) == max(nchar(longest)))][1] }
  
  # return the k and kmer with names associated with the vector
  return_values = c(k, kmer)
  names(return_values) = c("k", "kmer")
  return(return_values)
}

find_kmer_per_language = function(entry1, entry2, fun1, fun2) {
  
  # extract longest common substring based on different word sets output from fun1 and fun2
  # example fun1 = get_citation_forms, fun2 = get_guidewords
  longest_kmer_x = get_longest_kmer_from_set_of_words(entry1, entry2, fun1)
  longest_kmer_y = get_longest_kmer_from_set_of_words(entry1, entry2, fun2)
  
  # extract the longest kmer from each set of comparisons
  lengths = c(longest_kmer_x["k"], longest_kmer_y["k"])
  names(lengths) = c(longest_kmer_x["kmer"], longest_kmer_y["kmer"])
  
  # return the kmer that is the longest between these two sets
  names(which(lengths == max(as.numeric(lengths))))[1]
}
```

We now use these functions to identify entries that are in the same section and section breaks (breaks between sections). We define a section break as any line pair where overlap in guidewords + citation forms is zero AND the longest shared substring (k) is below a defined cutoff (our cutoff was 3). Not all entries will be assigned to a section using this method.

```{r}
def_section_breaks = function(df, cutoff) {
  df$section = NA
  sect_num = 1
  first_section_start = which(df$overlap > 0 | df$k >= cutoff)[1]
  df$section[first_section_start] = sect_num
  section_break = FALSE
  
  for (i in (first_section_start):nrow(df)) {
    if (df$overlap[i] > 0 | df$k[i] >= cutoff) {
      if (section_break == TRUE) { sect_num = sect_num + 1 }
      df$section[i] = sect_num
      section_break = FALSE
    } 
    else { 
      df$section[i] = NA
      section_break = TRUE
    }}
  df
}
```

After defining section breaks, we export all sections (with all members of each) to an output file. We also name each section based on the most common guidewords in that section.

```{r}
extract_sections = function(df, file) {
  # writes all entries present in each section to a file
  section_nums = unique(df$section[!is.na(df$section)])
  sections = data.frame(sapply(section_nums, function(x) x = character(max(table(df$section)) + 1)))
  
  for (i in section_nums) {
    # extract all of the entries that belong to that section (including duplicates)
    lines_a = df[which(df$section == i),]$line_a
    lines_b = df[which(df$section == i),]$line_b
    elements = c(lines_a, tail(lines_b, n = 1))
    # define number of empty rows for that section in output dataframe
    missing = max(table(df$section) + 1) - length(elements)
    # add that number of NAs to the end of the elements vector
    elements = c(elements, rep(NA, missing))
    # append entries belonging to that section to the ith column of the output dataframe
    sections[,i] = elements 
    # the guidewords from the first entry in a section is used to name the section
#    colnames(sections)[i] = paste(get_guidewords(sections[1,i]), collapse = "_")
    entries = unique(c(df[which(df$section == i),]$line_a, df[which(df$section == i),]$line_b))
    t = table(get_guidewords(entries))
    colnames(sections)[i] = paste(names(which(t == max(t))), collapse = "_")
     }
  write.csv(sections, file, row.names = FALSE, quote = FALSE)
}
```

The function below wraps around the previously defined functions, reading in an input file, initializing a new data frame to store results, finding section breaks and outputing sections and entry by entry comparisons to new files. 

```{r}
compare_entries = function(infile, comp, cutoff, outfile1, outfile2) {
#  name = strsplit(infile, "\\.")[[1]][1]
  df_composite = read.csv(infile, stringsAsFactors = FALSE)
  # get lines that belong to composite text
  df_composite = df_composite[which(grepl(comp, df_composite$id_text)),]
  # remove lines in df representing missing lines or sections
  empty_lines = which(df_composite$entry == "")
  if (length(empty_lines) != 0) {df_composite = df_composite[-empty_lines,] }
  
  # initialize empty data frame for storing results
  num_rows = nrow(df_composite) - 1
  df_compare = data.frame(line_a = character(num_rows), 
                          line_b = character(num_rows),
                          overlap = numeric(num_rows),
                          kmer = character(num_rows),
                          k = numeric(num_rows),
                          stringsAsFactors = FALSE)
  
  for (i in 1:num_rows) {
    line_a = tolower(df_composite$entry[i])
    guidewords_a = get_guidewords(line_a)
    citation_form_a = get_citation_forms(line_a)
    
    line_b = tolower(df_composite$entry[i + 1])
    guidewords_b = get_guidewords(line_b)
    citation_form_b = get_citation_forms(line_b)

  kmer = find_kmer_per_language(line_a, line_b, get_guidewords, get_citation_forms) 
  if (is.na(kmer)) k = 0 else k = nchar(kmer)
  
    overlap = sum(citation_form_a %in% citation_form_b) + sum(guidewords_a %in% guidewords_b)
    
    df_compare$line_a[i] = line_a
    df_compare$line_b[i] = line_b
    df_compare$overlap[i] = overlap
    df_compare$kmer[i] = kmer
    df_compare$k[i] = k
  }
#  plot(df_compare$overlap, pch = ".", main = name, ylab = "# matching words", xlab = "line number")
  df_compare = def_section_breaks(df_compare, cutoff = cutoff)
  write.csv(df_compare, outfile2, row.names = FALSE, quote = FALSE)
#  plot(table(df_compare$section), main = name, ylab = "# entries in section", xlab = "section")
  extract_sections(df_compare, outfile1)
  df_compare
}
```

```{r eval = TRUE}
Q1 = suppressWarnings(compare_entries(infile = "data/pass/Q01_par.csv", comp = "Q000001", cutoff = 3, outfile1 = "data/sections/Q01_sections.csv", outfile2 = "data/sections/intermediate/Q01_comparison_table.csv"))
Q39 = suppressWarnings(compare_entries(infile = "data/pass/Q39_par.csv", comp = "Q000039", cutoff = 3, outfile1 = "data/sections/Q39_sections.csv", outfile2 = "data/sections/intermediate/Q39_comparison_table.csv"))
Q40 = suppressWarnings(compare_entries(infile = "data/pass/Q40_par.csv", comp = "Q000040", cutoff = 3, outfile1 = "data/sections/Q40_sections.csv", outfile2 = "data/sections/intermediate/Q40_comparison_table.csv"))
Q41 = suppressWarnings(compare_entries(infile = "data/pass/Q41_par.csv", comp = "Q000041", cutoff = 3, outfile1 = "data/sections/Q41_sections.csv", outfile2 = "data/sections/intermediate/Q41_comparison_table.csv"))
Q42 = suppressWarnings(compare_entries(infile = "data/pass/Q42_par.csv", comp = "Q000042", cutoff = 3, outfile1 = "data/sections/Q42_sections.csv", outfile2 = "data/sections/intermediate/Q42_comparison_table.csv"))
```
